---
output: 
  pdf_document:
    fig_caption: true
    keep_tex: true
    latex_engine: pdflatex
    number_sections: true
header-includes:
- \usepackage[english]{babel}
- \usepackage{setspace}
- \usepackage{bm}
- \hyphenation{}
- \pagenumbering{arabic}
- \numberwithin{equation}{section}
- \newcommand{\E}{\mathbb{E}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\argmax}{\text{argmax}}
fontsize: 10pt
geometry: a4paper, left=30mm, right=30mm, top=20mm, bottom=20mm, includefoot
bibliography: references.bib
documentclass: article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(MASS)
library(mixtools)
library(plyr)
library(xtable)
```

<!-- Titlepage -->
\thispagestyle{empty}

\begin{raggedright}
  Freie Universität Berlin\\
  Chair of Statistics\\
  Location: Berlin\\
  Summer Term 2017\\
  Lecture: Einführung in die Bayes-Statistik\\
  Examiner: Dr. Florian Meinfelder
\end{raggedright}

\vspace*{\fill}
  \begin{center}
    \textbf{\Large{Program leave-one-out posterior predictive checking in R}}%
  \end{center}
\vfill

\begin{raggedleft}
  Johannes Brinkmann (), \href{mailto:jojo-brinkmann@gmx.de}{jojo-brinkmann@gmx.de}\\
  Carlo Michaelis (5043128), \href{mailto:carlo.michaelis@gmail.com}{carlo.michaelis@gmail.com}\\
  Max Reinhardt (), \href{mailto:maxreinhardt@me.com}{max\_reinhardt@me.com}\\
  Adrian Rolf (), \href{mailto:adrian.rolf@gmx.de}{adrian.rolf@gmx.de}\\
  \ \\
  Master Statistics\\
  \today\\
\end{raggedleft}

\newpage

<!-- Table of Contents -->
\tableofcontents
\newpage

<!-- List of Figures and Tables -->
\listoffigures
\listoftables
\newpage

<!-- Document Body -->
# Introduction

<!-- References -->
# References